Synthesising Recursive Functions for\\ First-Order Model Counting

### 2023-04-19 doc/talks/3_long/talk.tex
Minor changes based on feedback

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{18th April 2023}

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + mf(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming.
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}.
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a \alert{$\#\P$-complete} problem.
          \pause
    \item The rest of this talk is about how to construct such functions
          automatically.
  \end{itemize}
\end{frame}

@@

@@

        \node[below=0.3cm of simplification] (simplified) {\tiny $f(m, n) = f(m, n-1) + mf(m-1, n-1)$};

@@

@@

\begin{frame}{How to Interpret an FCG}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \vspace{1cm}
      \begin{forest}
        for tree={sn edges}
        [$\GDR$,ellipse,draw,name=gdr,fill=red!50,fill on=<4>
        [$\bigvee$,ellipse,draw,name=bigvee,fill=red!50,fill on=<5>
        [$\CR$,ellipse,draw,name=cr,fill=red!50,fill on=<6>
        [$\land$,ellipse,draw,name=conj,fill=red!50,fill on=<7>
        [$\bot$,rectangle,draw,alt=<8>{fill=red!50}{fill=gray!25},name=contradiction]
        [$\Reff$,ellipse,draw,name=ref,fill=red!50,fill on=<9>]
        ]
        ]
        ]
        ]
        \draw[-Latex,bend right=45] (ref) to (gdr);
        \draw[visible on=<2-3>] (5, 42 |- gdr) node[text width=5cm,name=gdrtwo,text=color2] {\alert<3>{Generalised domain recursion}};
        \draw[visible on=<2-3>] (5, 42 |- bigvee) node[text width=5cm,name=bigveetwo,text=color2] {Deterministic set-disjunction};
        \draw[visible on=<2-3>] (5, 42 |- cr) node[text width=5cm,name=crtwo,text=color2] {\alert<3>{Constraint removal}};
        \draw[visible on=<2-3>] (5, 42 |- conj) node[text width=5cm,name=conjtwo,text=color2] {Decomposable conjunction};
        \draw[visible on=<2-3>] (5, 42 |- ref) node[text width=5cm,name=reftwo,text=color2] {\alert<3>{Caching}};
        \node[visible on=<2-3>,below=0.5cm of contradiction,name=contradictiontwo,text=color2] {Contradiction};
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (gdrtwo) -- (gdr);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (bigveetwo) -- (bigvee);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (crtwo) -- (cr);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (conjtwo) -- (conj);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (reftwo) -- (ref);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (contradictiontwo) -- (contradiction);
      \end{forest}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{align*}
        \onslide<4->{f(m, n) &=} \onslide<5->{\sum_{l=0}^m \binom{m}{l}} \onslide<8->{[l<2]} \onslide<7->{\times} \onslide<9->{f(m-l, n-1)}\\
        \onslide<10>{&= f(m, n-1) + mf(m-1, n-1)}
      \end{align*}
      \onslide<8>{
        \[
          [\phi] =
          \begin{cases}
            1 & \text{if } \phi \\
            0 & \text{if } \neg\phi
          \end{cases}
          \]
      }
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{How These Rules Fit Together (2/5)}
  \begin{tikzpicture}[overlay,remember picture]
    \node[draw,rounded rectangle,fill=color2!50,anchor=center,below=1cm] at (current page.center) (rules) {Atom counting and unit propagation};
    \node[above=of rules] (before) {$\begin{gathered}
        \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(\alert{c}, z)\\
        \forall x \in \Gamma\text{. }\forall y,z \in \Delta\text{. }y \ne z \land \alert{x \ne c} \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{x \ne c} \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(x, y)\\
        \forall w \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(\alert{c}, y)\\
        \forall w,x \in \Gamma\text{. }\forall y \in \Delta\text{. }w \ne x \land \alert{w \ne c} \land \alert{x \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \node[below=of rules] (after) {$\begin{gathered}
        \forall y, z \in \alert{\Delta^{\top}}\text{. }y \ne z \Rightarrow \bot\\
        \forall x \in \Gamma\text{. }\forall y, z \in \alert{\Delta^{\bot}}\text{. }y \ne z \land x \ne c \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z) \\
        \forall w, x \in \Gamma\text{. }\forall y \in \alert{\Delta^{\bot}}\text{. }w \ne x \land w \ne c \land x \ne c \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (before) -- (rules);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
  \end{tikzpicture}
\end{frame}

>>>>

\author{Paulius Dilkas \\ \vspace{1cm} \tiny{Joint work with Vaishak Belle (Univeristy of Edinburgh, UK)}}
\title{Synthesising Recursive Functions for\\ First-Order Model Counting}
\institute{National University of Singapore, Singapore}
\date{PLSE Seminar, 21st April 2023}

@@

\begin{frame}{Back to Our Example}
  The following function counts injections:
  \[
  f(m, n) =
  \begin{cases}
    1 & \text{if } m = 0 \text{ and } n = 0 \\
    0 & \text{if } m > 0 \text{ and } n = 0 \\
    f(m, n-1) + m \times f(m-1, n-1) & \text{otherwise.}
  \end{cases}
  \]
  \pause
  \begin{itemize}
    \item \alert{$f(m, n)$} can be computed in \alert{$\Theta(mn)$} time
    \begin{itemize}
      \item using dynamic programming.
    \end{itemize}\pause
    \item \alert{Optimal} time complexity to compute \alert{$n^{\underline{m}}$}
          is \alert{$\Theta(m)$}.
    \item But \alert{$\Theta(mn)$} is still much better than translating to
          propositional logic and solving a \alert{$\#\P$-complete} problem.
          \pause
    \item The rest of this talk is about how to construct such functions
          automatically.
  \end{itemize}
\end{frame}

@@

        \node[draw,fit={(kc) (graph) (evaluation)},inner xsep=1pt,inner ysep=2pt] {};

@@

        \node[below=0.3cm of simplification] (simplified) {\tiny $f(m, n) = f(m, n-1) + m \times f(m-1, n-1)$};

@@

        \node[draw,fit={(kc) (graph) (interpretation)},inner xsep=1pt,inner ysep=2pt,visible on=<3->] {};

@@

\begin{frame}{How to Interpret an FCG}
  \begin{columns}
    \begin{column}{0.25\textwidth}
      \centering
      \vspace{1cm}
      \begin{forest}
        for tree={sn edges}
        [$\GDR$,ellipse,draw,name=gdr,fill=red!50,fill on=<4>
        [$\bigvee$,ellipse,draw,name=bigvee,fill=red!50,fill on=<5>
        [$\CR$,ellipse,draw,name=cr,fill=red!50,fill on=<6>
        [$\land$,ellipse,draw,name=conj,fill=red!50,fill on=<7>
        [$\bot$,rectangle,draw,alt=<8>{fill=red!50}{fill=gray!25},name=contradiction]
        [$\Reff$,ellipse,draw,name=ref,fill=red!50,fill on=<9>]
        ]
        ]
        ]
        ]
        \draw[-Latex,bend right=45] (ref) to (gdr);
        \draw[visible on=<2-3>] (5, 42 |- gdr) node[text width=5cm,name=gdrtwo,text=color2] {\alert<3>{Generalised domain recursion}};
        \draw[visible on=<2-3>] (5, 42 |- bigvee) node[text width=5cm,name=bigveetwo,text=color2] {Deterministic set-disjunction};
        \draw[visible on=<2-3>] (5, 42 |- cr) node[text width=5cm,name=crtwo,text=color2] {\alert<3>{Constraint removal}};
        \draw[visible on=<2-3>] (5, 42 |- conj) node[text width=5cm,name=conjtwo,text=color2] {Decomposable conjunction};
        \draw[visible on=<2-3>] (5, 42 |- ref) node[text width=5cm,name=reftwo,text=color2] {\alert<3>{Caching}};
        \node[visible on=<2-3>,below=0.5cm of contradiction,name=contradictiontwo,text=color2] {Contradiction};
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (gdrtwo) -- (gdr);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (bigveetwo) -- (bigvee);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (crtwo) -- (cr);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (conjtwo) -- (conj);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (reftwo) -- (ref);
        \draw[visible on=<2-3>,-Latex,dashed,color=color2] (contradictiontwo) -- (contradiction);
      \end{forest}
    \end{column}
    \begin{column}{0.75\textwidth}
      \begin{align*}
        \onslide<4->{f(m, n) &=} \onslide<5->{\sum_{l=0}^m \binom{m}{l}} \onslide<8->{[l<2]} \onslide<7->{\times} \onslide<9->{f(m-l, n-1)}\\
        \onslide<10->{&= \binom{m}{0} \times f(m-0, n-1)}\\
        \onslide<10->{&+ \binom{m}{1} \times f(m-1, n-1)}\\
        \onslide<11->{&= f(m, n-1) + m \times f(m-1, n-1)}
      \end{align*}
      \onslide<8>{
        \[
          [\phi] =
          \begin{cases}
            1 & \text{if } \phi \\
            0 & \text{if } \neg\phi
          \end{cases}
          \]
      }
    \end{column}
  \end{columns}
\end{frame}

@@

\begin{frame}{How These Rules Fit Together (2/5)}
  \begin{tikzpicture}[overlay,remember picture]
    \node[draw,rounded rectangle,fill=color2!50,anchor=center,below=1cm] at (current page.center) (rules) {Atom counting and unit propagation};
    \node[above=of rules] (before) {$\begin{gathered}
        \forall y, z \in \Delta\text{. }y \ne z \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(\alert{c}, z)\\
        \forall x \in \Gamma\text{. }\forall y,z \in \Delta\text{. }y \ne z \land \alert{x \ne c} \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z)\\
        \forall x \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{x \ne c} \Rightarrow \neg\pP(\alert{c}, y) \lor \neg\pP(x, y)\\
        \forall w \in \Gamma\text{. }\forall y \in \Delta\text{. }\alert{w \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(\alert{c}, y)\\
        \forall w,x \in \Gamma\text{. }\forall y \in \Delta\text{. }w \ne x \land \alert{w \ne c} \land \alert{x \ne c} \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \node[below=of rules] (after) {$\begin{gathered}
        \forall y, z \in \alert{\Delta^{\top}}\text{. }y \ne z \Rightarrow \bot\\
        \forall x \in \Gamma\text{. }\forall y, z \in \alert{\Delta^{\bot}}\text{. }y \ne z \land x \ne c \Rightarrow \neg\pP(x, y) \lor \neg\pP(x, z) \\
        \forall w, x \in \Gamma\text{. }\forall y \in \alert{\Delta^{\bot}}\text{. }w \ne x \land w \ne c \land x \ne c \Rightarrow \neg\pP(w, y) \lor \neg\pP(x, y)
      \end{gathered}$};
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (before) -- (rules);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
    \draw[-Latex,color=color2,ultra thick,shorten <=0.1cm,shorten >=0.1cm] (rules) -- (after);
  \end{tikzpicture}
\end{frame}
% NOTE: need to explain how atom counting splits the domain
